COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-23 TACO-paper/coffee.tex
Elaborate padding. Minor fixes

% TODO: could add a sentence here to bridge the two paragraphs

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g. AVX processor, with 256 bit long vector registers and 64-bit double-precision floating-point values). In this case, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers exhibit. First, they only identify sub-expressions that are invariant with respect to the innermost loop. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could otherwise be significant. It is also worth noting that, in some problems (e.g. in Helmholtz), invariant sub-expressions along $j$ are identical to those along $k$, and both loops iterate over the same iteration space. In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and the variables accessed are such that the vector-register pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Standard optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,3
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. In addition to incurring in larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
\caption{Outer-product vectorization by permuting values in a vector register.}
\label{fig:vect-by-vect}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.60]{Pictures/vect-restore-inline.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsic.}
\label{fig:restore-layout}
\end{figure}

@@

\begin{algorithm}[t]
\footnotesize
  \textbf{The COFFEE Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)\\
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
// Optimize ast based on plan \\
ast.licm() \\
ast.padding() \\
ast.data\_align() \\
\If{plan.permute}{
  ast.permute\_assembly\_loops()
}
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  uaj = MIN(plan.uaj\_factor, $\lfloor$it\_space.j.size/isa.vf$\rfloor$) \\
  ast.op\_vect(uaj)
}
\Return{wrapper + ast.from\_ast\_to\_c()}
\caption{Pseudocode of the COFFEE pipeline.}
\label{algo:PyOP2Compiler}
\end{algorithm}

>>>>

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
The absence of stencils renders the element matrix computation easily auto-vectorizable by a vendor compiler. Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are particularly small as in the case of assembly.

Data alignment is enforced in two steps. Firstly, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g. AVX processor, with 256 bit long vector registers and 64-bit double-precision floats). In this case, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma.

Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient. These extra iterations only write to the padded region of the element matrix, and therefore have no side effects on the final result.

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers exhibit. First, they only identify sub-expressions that are invariant with respect to the innermost loop. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization.

We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could otherwise be significant. It is also worth noting that, in some problems (e.g. in Helmholtz), invariant sub-expressions along $j$ are identical to those along $k$, and both loops iterate over the same iteration space. In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and the variables accessed are such that the vector-register pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Standard optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,3
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,3$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. In addition to incurring in larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.8]{Pictures/vect-by-vect-inline.pdf}}
\caption{Outer-product vectorization by permuting values in a vector register.}
\label{fig:vect-by-vect}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[scale=0.8]{Pictures/vect-restore-inline.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsic.}
\label{fig:restore-layout}
\end{figure}

@@

\begin{algorithm}[t]
\footnotesize
  \textbf{The COFFEE Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)\\
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
// Optimize ast based on plan \\
ast.licm() \\
ast.padding() \\
ast.data\_align() \\
\If{plan.permute}{
  ast.permute\_assembly\_loops()
}
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  uaj = MIN(plan.uaj\_factor, $\lfloor$it\_space.j.size/isa.vf$\rfloor$) \\
  ast.op\_vect(uaj)
}
\Return{wrapper + ast.from\_ast\_to\_c()}
\caption{Pseudocode of the COFFEE architecture.}
\label{algo:PyOP2Compiler}
\end{algorithm}
